High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-05-30 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

The resulting {\tt uvmem} image shown in the top right panel of Fig.~\ref{fig:images_vs_simulated} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, beyond which it drops steeply before breaking at $\sim$35\,au into a shallow tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} did anticipate its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au. The {\tt uvmem} map has a resolution of 21$\times$18\,mas and an RMS noise of 26$\micron$Jy per resolution beam, at a frequency of 237 GHz.

@@

The ring width can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. The top panel of Fig.~\ref{fig:polarring} shows the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim$0$\farcs021 \,\times \,\sim$0$\farcs018$, we see that Ring13 is marginally resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

@@

Interestingly, the ALMA image also detects faint 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this central emission is larger than the angular resolution, it is probably stemming from thermal emission from large dust grains rather than directly from the stars. The peak intensity of this dust structure is at an estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au, from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au. \citet{Francis_2020} also did some measurements of this central emission by assuming a power--law profile for an inner disc that goes from 0.1 to 4\,au, and has a total dust mass of 0.013$\pm$0.002\,M$_{\earth}$.

@@

The system present a low near--infrared (NIR) excess of 0.9$\pm$3.7\,\% \citep{Francis_2020}. This appears to be inconsistent with the detection of the mm-bright central emission, but the low NIR excess could be interpreted as a lack of micron-sized dust in the central blob. Also, from this feature (or the absence of it) \citet{Francis_2020} stated that the inner radius of the millimetre-sized dust ring that accounts for the central emission is well beyond the sublimation radius.

@@

For the vertical structure, \citet{dOrazi} found flaring angles (i.e. height of the ring over the disc midplane divided by the radius of the ring), of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values was set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is set to $H_\circ = 0.94$\,au at $r_\circ = 18$\,au. This vertical structure is consistent with the measurements made by \citet{dOrazi}, as the model flaring angles are $\varphi_{\mathrm{inner}}$ = 6.5\,$\degr$ and $\varphi_{\mathrm{outer}}$ = 7.9\,$\degr$. This is summarized in bottom panel of Fig.~\ref{fig:profiles} that shows the aspect ratio profile $h(r)=H(r)/r$.

@@

We chose to use Gaussian profiles to parametrize Ring5 and Ring13,
\begin{equation}
  \Sigma_{\mathrm{ring}}(r) = \frac{\Sigma_\circ}{\sqrt{2 \pi} \sigma}
  \, \mathrm{exp}\left[-\frac{1}{2}\left(\frac{r-\mu}{\sigma}\right)^{2}\right],
\end{equation}
where we define constants that correspond to the centroid radii \{$\mu_{\mathrm{Ring5}}=5.2$\,au, $\mu_{\mathrm{Ring13}}=14.9$\,au\}, ring widths \{$\sigma_{\mathrm{Ring5}}=0.25$\,au, $\sigma_{\mathrm{Ring13}}=2.2$\,au\}, and normalizations \{$\Sigma_{\circ,\mathrm{Ring5}}=3.3$\,$\mathrm{g\,cm^{-3}}$, $\Sigma_{\circ,\mathrm{Ring13}}=6.0 \times 10^{-1}$\,$\mathrm{g\,cm^{-3}}$\} for both components separately.

@@

As mentioned before, the model presented here gives a solution to a highly degenerative problem. Nevertheless, as a way to improve and obtain a rough measure of the accuracy of our model, we made one dimensional explorations of the parameter space and found uncertainties of some relevant parameters that will be useful for the discussion (see Appendix.~\ref{sec:Appendix}). We can estimate confidence intervals for the scale height at $r_\circ$= 18\,au with $H_\circ = 0.92^{+0.01}_{-0.01}$\,au, for the location of Ring5 with $r=5.2^{+0.14}_{-0.36}$\,au, and for the width of gas Ring13 of $w_\mathrm{g} = 5.18^{+0.33}_{-0.20}$\,au.

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. The introduced Ring5 is not visible in the simulated image as its predicted peak intensity is around two times the noise in the ALMA image ($\sim$1$\times10^{-7}$\,Jy\,$\mathrm{beam}^{-1}$). This gives us an upper limit for the total millimetre--sized dust mass present in Ring5 of $\sim$2$\times10^{-5}$\,M$_{\earth}$. The depletion of large dust grains in Ring5 is consistent with efficient dust trapping in Ring13. A zone of radially increasing gas pressure can entirely filter out large dust grains from inner regions, while smaller grains can be able to overcome this barrier due to their strong frictional coupling to the gas accretion flow \citep[studied in the context of planetary gaps,][]{Rice2006,Zhu2012,Weber2018}. Still, the strong depletion of large particles demands that within Ring5 dust evolution is extremely inefficient or limited, or that large particles are rapidly removed. Otherwise, the present small grains should quickly coagulate to form detectable grain sizes in the ALMA observation in Fig.~\ref{fig:images_vs_simulated} \citep{Drazkowska2019}. The central emission is also well reproduced with the inclusion of a Gaussian close ring, resulting in the observed central blob. The lack of a high NIR excess in the SED indicate that the dust in this blob must be at least bigger than 0.5\,mm (see Appendix~\ref{sec:Appendix}), but we choose to use 0.8\,mm as the lower limit for the dust population that composes this feature because this predicts a dust mass of 0.012\,M$_{\earth}$, that is in agreement with the mass estimate of \citet{Francis_2020} (0.013$\pm$0.002\,M$_{\earth}$).

@@

Discs, rings and gaps are common structures present in protoplanetary discs, each time a disc with complex geometry appears we learn something knew about planet formation or disc mechanics.

Observing the same source at different wavelengths, or employing more than one instrument, will tell different stories about the same physical phenomena. Using radiative transfer simulations let us find a common plot that may explain all this diverse results. For V4046\,Sgr, there was a missing piece for an accurate representation of the system: a thin ring of millimetre-sized dust grains at a distance of $\sim$13\,au. We present new ALMA 1.3\,mm continuum imaging of this circumbinary disc at an unprecedented definition ($\sim$0$\farcs021 \, \times \, \sim$0$\farcs018$) were new features ring becomes visible for the first time. The new observed feature gives us some insights of the dynamics that are driving the formation of this unique morphology. Together with the analysis of SPHERE--IRDIS polarized images and a well-sampled SED, the radiative transfer modeling came as a useful tool for increasing the scope of our investigation.

@@

  \item The central emission present in the ALMA image might be a close inner ring made up of dust grains larger than 0.5\,mm, or it would create a NIR excess higher than the one inferred from the SED. Our model agrees with the mass estimation for this feature made by \citet{Francis_2020} with a mass of 0.012\,M$_{\earth}$.

@@

  \item Our parametric model, which accounts for the SED of the system, involves the existence of a sub-micron dust population close to the stars (inner disc). Also predicts the presence of another inner ring at $\sim$5\,au, mainly consisting of small dust grains. The depletion of large dust is consistent with efficient dust trapping in Ring13. This additional ring lies under the coronagraph of the scattered light image and is to faint to be detected by the ALMA observation.
\end{enumerate}

>>>>

The resulting {\tt uvmem} image shown in the top right panel of Fig.~\ref{fig:images_vs_simulated} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, continues relatively flat and then breaks at $\sim$35\,au into a steeper tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} did anticipate its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au. The {\tt uvmem} map has a resolution of 21$\times$18\,mas and an RMS noise of 26$\micron$Jy per resolution beam, at a frequency of 237 GHz.

@@

The ring width can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. The top panel of Fig.~\ref{fig:polarring} shows the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim$0$\farcs021 \,\times \,\sim$0$\farcs018$ (or $\sim$1.4\,au at 71.1\,pc), we see that Ring13 is resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

@@

Interestingly, the ALMA image also detects 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this central emission is larger than the angular resolution, it is probably stemming from thermal emission from large dust grains rather than directly from the stars. The peak intensity of this dust structure is at an estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au, from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au. Measurements of this central emission made by \citet{Francis_2020}, assume a power-law profile for an inner disc that goes from 0.1 to 4\,au, and has a total dust mass of 0.013$\pm$0.002\,M$_{\earth}$.

@@

The SED present a low near--infrared (NIR) excess of 0.9$\pm$3.7\,\% \citep{Francis_2020}. This appears to be inconsistent with the detection of the mm-bright central emission, as the presence of a inner disc would generate higher NIR fluxes, but the low excess could be interpreted as a lack of micron-sized dust in the central spot. Also, from this feature (or the absence of it) \citet{Francis_2020} stated that the inner radius of the millimetre-sized dust ring that accounts for the central emission is well beyond the sublimation radius, around 0.1\,au.

@@

For the vertical structure, \citet{dOrazi} found flaring angles (i.e. height of the ring over the disc midplane divided by the radius of the ring), of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values was set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is set to $H_\circ = 0.91$\,au at $r_\circ = 18$\,au. This vertical structure is consistent with the measurements made by \citet{dOrazi}, as the model flaring angles are $\varphi_{\mathrm{inner}}$ = 6.5\,$\degr$ and $\varphi_{\mathrm{outer}}$ = 7.8\,$\degr$. This is summarized in bottom panel of Fig.~\ref{fig:profiles} that shows the aspect ratio profile $h(r)=H(r)/r$.

@@

We chose to use Gaussian profiles to parametrize Ring5 and Ring13,
\begin{equation}
  \Sigma_{\mathrm{ring}}(r) = \frac{\Sigma_\circ}{\sqrt{2 \pi} \sigma}
  \, \mathrm{exp}\left[-\frac{1}{2}\left(\frac{r-\mu}{\sigma}\right)^{2}\right],
\end{equation}
where we define constants that correspond to the centroid radii \{$\mu_{\mathrm{Ring5}}=5.2$\,au, $\mu_{\mathrm{Ring13}}=14.9$\,au\}, ring widths \{$\sigma_{\mathrm{Ring5}}=0.25$\,au, $\sigma_{\mathrm{Ring13}}=2.25$\,au\}, and normalizations \{$\Sigma_{\circ,\mathrm{Ring5}}=3.3$\,$\mathrm{g\,cm^{-3}}$, $\Sigma_{\circ,\mathrm{Ring13}}=6.0 \times 10^{-1}$\,$\mathrm{g\,cm^{-3}}$\} for both components separately.

@@

As mentioned before, the model presented here gives a solution to a highly degenerative problem. Nevertheless, as a way to improve and obtain a rough measure of the accuracy of our model, we made one dimensional explorations of the parameter space and found uncertainties of some relevant parameters that will be useful for the discussion (see Appendix.~\ref{sec:Appendix}). We can estimate confidence intervals for the scale height at $r_\circ$= 18\,au with $H_\circ = 0.91^{+0.01}_{-0.01}$\,au and for the width of gas Ring13 of $w_\mathrm{g} = 5.30^{+0.24}_{-0.30}$\,au.

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. The introduced Ring5 is not visible in the simulated image as its predicted peak intensity is around two times the noise in the ALMA image ($\sim$1$\times10^{-7}$\,Jy\,$\mathrm{beam}^{-1}$). This gives us an upper limit for the total millimetre--sized dust mass present in Ring5 of $\sim$2$\times10^{-5}$\,M$_{\earth}$. The depletion of large dust grains in Ring5 is consistent with efficient dust trapping in Ring13. A zone of radially increasing gas pressure can entirely filter out large dust grains from inner regions, while smaller grains can be able to overcome this barrier due to their strong frictional coupling to the gas accretion flow \citep[studied in the context of planetary gaps,][]{Rice2006,Zhu2012,Weber2018}. Still, the strong depletion of large particles demands that within Ring5 dust evolution is extremely inefficient or limited, or that large particles are rapidly removed. Otherwise, the present small grains should quickly coagulate to form detectable grain sizes in the ALMA observation in Fig.~\ref{fig:images_vs_simulated} \citep{Drazkowska2019}. The central emission is also well reproduced with the inclusion of a Gaussian close ring. The lack of a high NIR excess in the SED indicate that the dust in this blob must be at least bigger than 0.5\,mm (see Appendix~\ref{sec:Appendix}), but we choose to use 0.8\,mm as the lower limit for the dust population that composes this feature because this predicts a dust mass of 0.012\,M$_{\earth}$, that is in agreement with the mass estimate of \citet{Francis_2020} (0.013$\pm$0.002\,M$_{\earth}$).

@@

Rings, spirals, cavities and gaps are common structures present in protoplanetary discs, but only recent observations have allowed us to study them in detail. Observing systems with solar-type stars, that are not exceptionally bright or large, has not been the main focus of high resolution imaging. For that reason, DARTTS-S programme, that collected DPI data for T Tauri Stars, has been so revealing, and with the counterpart of ALMA continuum maps from DARTTS-A, we can search for substructures in this systems as never before. Following that line, we present new ALMA 1.3\,mm continuum imaging of V4046\,Sgr, a well know circumbinary disc, at an unprecedented definition ($\sim$0$\farcs021 \, \times \, \sim$0$\farcs018$) where new features become visible. Together with the analysis of a SPHERE--IRDIS polarized image and a well-sampled SED, we aim to reproduce the system with radiative transfer modeling looking for a way to explain the data in terms of a physically model.

@@

  \item The central emission present in the ALMA image suggests the existence of a close inner ring that goes in pass the sublimation radius and is made up of dust grains larger than 0.5\,mm. Our interpretation agrees with the mass estimation for this feature made by \citet{Francis_2020} with a mass of 0.012\,M$_{\earth}$.

  \item Our parametric model, which accounts for the SED of the system, involves the existence of sub-micron dust population close to the stars (inner disc outside the sublimation radius). Also predicts the presence of another inner ring at $\sim$5\,au, mainly consisting of small dust grains. The depletion of large dust is consistent with efficient dust trapping in Ring13. This additional ring lies under the coronagraph of the scattered light image and is to faint to be detected by the ALMA observation.

@@

\end{enumerate}
